### 2021-11-15 paper/output/document/whykant.tex
made writing fixes

%
\begin{isabellebody}%
\setisabellecontext{whykant}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Why Kantian Ethics%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this thesis, I automate Kantian ethics. In 2006, Powers posited that deontological theories are 
attractive candidates for automation because rules are generally computationally tractable \cite[1]{powers}. 
Intuitively, algorithms are rules or procedures for problem solving and Kantian ethics offer one such 
procedure for the problem of making ethical judgements. I will make this intuition precise by
arguing that Kantian ethics is natural to formalize because evaluating a maxim requires little additional
data about the world and a maxim is relatively easy to represent to a computer. All 
ethical traditions have debates that an automated ethical system will need to take a stance on, but these debates
are less frequent and less controversial for Kantian ethics than for consequentialism and 
virtue ethics.

I do not aim to show that Kantian ethics is the only tractable theory to automate and
to present a comprehensive overview of all consequentialist or virtue ethical theories. Instead, I 
present a sample of some approaches in each tradition and argue that Kantian ethics is more straightforward 
to formalize than these approaches. Future work could and should address the challenges I outline in 
this section. The more ethical theories that computational tools can handle, the more valuable 
computational philosophy becomes both for philosophers and for AI agents. Insofar as my project serves 
as an early proof-of-concept for computational ethics, I choose to automate an ethical theory that 
poses fewer challenges than others.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Kantian Ethics%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
\textbf{Crash Course on Kantian Ethics}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Kant's theory is centered on practical reason, which is the kind of reason that we 
use to decide what to do. In \emph{The Groundwork of the Metaphysics of Morals}, Kant's most influential 
text on ethics, he explains that rational beings are unique because we can act ``in accordance with 
the representations of laws" \cite[4:412]{groundwork}. In contrast, a ball thrown into the air acts 
according to the laws of physics themselves. It cannot ask itself, ``Should I fall back to the ground?" 
It simply falls. A rational being on the other hand, can ask themselves ``Should I act on this reason?" 
As Korsgaard describes it, when choosing which desire to act on, ``it is as if there is something over 
and above all of your desires, something which is you, and which chooses which desire to act on" \cite[100]{sources}. 
Rational beings are set apart by this reflective capacity. A rational being's behavior is purposive and 
their actions are guided by practical reason. They have reasons for acting, even when these reasons may be 
opaque to them. This operation of practical reason is what Kant calls the will.

The will operates by adopting, or willing, maxims, which are its percieved reasons for acting. Kant defines a maxim as 
the ``subjective principle of willing," or the reason that the will \emph{subjectively} gives 
to itself for acting \cite[16 footnote 1]{groundwork}. There is debate about what exactly must be 
included in a maxim, but most philosophers agree that a maxim consists of some combination of circumstances, 
act, and goal.\footnote{For more discussion of the definition of a maxim, see Section What Is a Maxim}
One example of a maxim is ``when I am hungry, I will eat a doughnut in order to satisfy my sweet tooth." 
When an agent wills this maxim, they decide to act on it. They commit themselves to the end in the maxim 
(e.g. satisfying your sweet tooth). They represent their action, to themselves, as following the 
principle given by this maxim. Because a maxim captures an agent's principle of action, Kant evaluates
maxims are obligatory or prohibited. He argues that certain maxims have a form or logical structure 
that requires any rational agent to will them, and these maxims are obligatory.

The form of an obligatory maxim is given by the categorical imperative, or the supreme rule of morality. 
An imperative is a command, such as ``Close the door" or "Eat the doughnut in order to satisfy your 
sweet tooth." An imperative is categorical if it holds unconditionally for all rational agents under all 
circumstances. Kant argues that the moral law must be a categorical imperative, for otherwise it would 
not have the force that makes it a moral law \cite[5]{groundwork}. In order for an imperative to be 
categorical, it must be derived from the will's authority over itself. Our wills are autonomous, so 
the only thing that can have unconditional authority over a rational will is 
the rational will itself. In Velleman's version of this argument, he claims that no one else can tell you what 
to do because you can always ask why you 
should obey their authority. The only authority that you cannot question is the authority of your own 
practical reason. To question this authority is to demand a reason for acting for reasons, which 
concedes the authority of reason itself \cite[23]{velleman}. Therefore, the only possible candidates 
for the categorical imperative are those rules that are required of the will because it is a will. 
The categorical imperative must be a property of practical reason itself.

Armed with this understanding of practical reason, Kant presents the categorical 
imperative. He presents three ``formulations" or versions of the categorical imperative and goes on to 
argue that all three formulations are equivalent. In this project, I focus on the first formulation,
the Formula of Universal Law, but will briefly present the other two as well\footnote{For more on this 
choice, see Section Why FUL}.

The above is not meant to serve as a full defense of Kant's ethical theory, as that is outside the scope
of this thesis. Instead, I briefly reconstruct an argument for Kant's ethical theory in the hopes 
of offering context for the implementation of the FUL I present later in the thesis. Additionally, understanding 
the structure of Kant's argument also reveals facts about his theory that make it an ideal candidate 
for formalization.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Ease of Automation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Kantian ethics is natural to formalize because the categorical imperative, particularly the FUL, 
is a formal principle of practical reason. In other words, the FUL is a property of reason related 
to the form or structure of a maxim. It has nothing to do with the circumstances of behavior 
(beyond those included in a maxim), the agent's mental state, or any other contingent facts. Instead, 
it is purely a property of the proposed principle for action. This formalism makes Kantian ethics an 
attractive candidate for formalization. While other ethical theories often rely on many facts about 
the world or the actor, Kantian ethics simply relies on the form of a given maxim. A computer evaluating 
a maxim doesn't require any knowledge about the world beyond what is contained in a maxim. A maxim 
is the only input that the computer needs to make a moral judgement. Automating 
Kantian ethics merely requires making the notion of a maxim precise and representing it to the computer. 
This distinguishes Kantian ethics from consequentialism and virtue ethics, which, as I will argue below, 
require far more knowledge about the world or the agent to reach a moral decision.

Not only does evaluating Kantian ethics only require evaluating a maxim, a maxim itself is an object
with a thin representation for a computer, as contrasted to more complex objects like states of 
affairs or moral character. Later in my project, I argue that a maxim can be represented simply as 
a tuple of circumstances, act, and goal\footnote{For more, see Section What is a Maxim?}. This representation
is simple and efficient, especially when compared to the representation of a causal chain or a state of 
affairs or moral character. A maxim is a principle with a well-defined form, so representing a maxim
to the computer merely requires capturing this form. This property not only reduces the computational complexity
(in terms of time and space) of representing a maxim, it also make the system easier for human reasoners
to interact with. A person crafting an input to a Kantian automated agent needs to reason about relatively
simple units of evaluation, as opposed to the more complex features that consequentialism and virtue
ethics reqire. I will make the comparision to consequentialism and virtue ethics explicit below.

I do not argue that Kantian ethics is the only theory possible to formalize, but instead that Kantian
ethics is relatively easier to formalize. Indeed, like any ethical theory, the Kantian tradition contains 
debates that philosophers disagree on. In this project, I assume stances on debates about
what exactly constitutes the form of a maxim and the correct interpretation of the Formula of 
Universal Law. Those who disagree with my stances will not trust the judgements of my system. Unlike
consequentialism or virtue ethics, these debates are considered close to settled in the Kantian literature 
\cite{ebelsduggan}. Because such debates are fewer and less controversial than those in virtue ethics
or consequentialism, Kantian ethics is relatively easier to formalize.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Consequentialism%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
A consequentialist ethical theory is, broadly speaking, any ethical theory that evaluates an action by evaluating 
its consequences. \footnote{There is long debate about what exactly makes an ethical theory consequentialist \cite{consequentialismsep}. 
For this paper, I will focus on theories that place the moral worth of an act in its the consequences.} For example, 
utilitarianism is a form of consequentialism in which the moral action 
is the action that results in the best consequences or produces the most good \cite{utilsep}. The focus
on the consequences of action distinguishes consequentialists from Kantians, who derive the moral worth
of an action from the maxim that is acted on. Some debates in the consequentialist tradition include 
which consequences of an action matter, what exactly constitutes a ``good" consequence, and how we can 
aggregate the consequences of an action over all the individuals involved.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Which Consequences Matter}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Because consequentialism evaluates the state of affairs following an action, this kind of ethical 
reasoning requires more knowledge
about the state of the world than Kantian ethics. Under a naive version of consequentialism, evaluating 
an action requires perfect knowledge of all consequences following an action. This requires that an 
automated ethical system somehow collect all of the infinite consquences following an action, a likely
impossible task. Moreover, compiling this database of consequences requires 
answering difficult questions about which consequences were actually caused by an action.\footnote
{maybe cite the debate about difficulties in determining causation?}

These challenges also apply to human reasoners, so most consequentialists do not actually adopt the naivie
view that agents need to calculate all the consequences of their actions. Plausible strategies to avoid 
this problem include stopping calculation early because constant calculation paralyzes action or only 
evaluating consequences that the agent could reasonably forsee before acting. Another solution is the 
``proximate cause" approach, which only holds the agent responsible for the immediate consequences of their 
acts, but not for consequences resulting from others' voluntary responses to the agent's original act \cite{consequentialismsep}.

Even without understanding the details of these views, it is clear that they require 
more data than Kantian ethics and scale poorly with the complexity of the act being evaluated. Even 
if we cut off the chain of causal reasoning at some point based on 
one of the rules above, evaluating the consequences of an action is still data-intensive. Even evaluating
the first or immediate cause of an action requires knowledge about the state of the world before
and after an action. Kantian ethics, on the other hand, requires only knowledge about the maxim, not 
knowledge about the state of the world when the maxim is adopted. Consequentialism requires knowledge about 
the situation in which the act is performed and following the act, whereas Kantian ethics merely requires 
knowledge about the act itself. For simple acts, collecting this data may not seem unreasonable, but as acts become
more complex and affect more people, the computational time and space required to calculate and store
their consequences increases. Kantian ethics, on the other hand, does not suffer this scaling
challenge because maxims that affect 1 person and maxims that affect 1 million people share the same
representation.

The fact that consequentialism requires more knowledge about the world makes it more difficult to formalize.
Automated consequentialist ethical systems would need to represent complex states of the world and causal chains
in an efficient manner and reason about them. This both presents a difficult technical challenge and
impedes the usability of such a system. Such a system would need to come equipped with 
a large enough database of knowledge about the world to extrapolate the consequences of an actions and
up-to-date information about the state of the world at the moment of action. Not only does collecting 
and representing this data pose a technical challenge, it also creates a larger ``trusted code base" for 
the automated system. Trusting my Kantian ethical reasoner merely requires trusting the logical implementation
of the categorical imperative and the formulation of a maxim, but trusting a consequentialist ethical 
reasoner requires trusting both the logical machinery that actually evaluates the act and the 
background/situational knowledge that serves as an input to this machinery.

The challenge of understanding and representing the circumstances of action is not unique to consequentialism,
but is particularly acute for consequentialism. Kantian ethicists robustly debate which circumstances 
of an action are ``morally relevant" and should be included in the formulation of the maxim.\footnote{Powers 
\cite{powers} identifies this as a challenge for automating Kantian ethics and briefly sketches 
solutions from O'Neill \citet{constofreason}, Silber \citet{silber}, and Rawls \citet{rawlsconstructivism}. } Those 
using my system will need to automatically or manually formulate maxims, a process that involves 
common-sense reasoning to determine which circumstances, act, and goal tuples the agent is adopting.
\footnote{For more on the challenge of parsing of ethical dilemmas into maxims, see Section AI Ethics}
Because Kantian ethics merely evaluates a single maxim, the surface of this debate is much smaller
than the debate about circumstances and consequences in a consequentialist system. An automated 
consequentialist system must make such judgements about the act itself, the circumstances in which 
it is performed, and the circumstances following the act. All ethical theories relativize
their judgements to the situation in which an act is performed to some extent, but consequentialism
requires far more knowledge about the world than Kantian ethics.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Theory of the Good}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Another debate that an automated consequentialist reasoner would need to take a stance on is
the question of what qualifies as a ``good consequence," or what the theory of the good is. Hedonists associate
good with the presence of pleasure and the absence of pain. Preference utiliarians believe that good is 
the satisfaction of desires and is thus derived from individuals' preferences, as opposed to some
sensation of pleasure or pain. Other consequentialists adopt a pluralistic theory of value, under which 
many different kinds of things are good for different reasons. For example, Moore values beauty and truth 
and other pluralists value justice, love, and freedom \cite{moorepe}. Welfare utilitarians value a person's 
welfare and utilitarians of right value states of affairs in which respect for some set of
rights is maximized \cite{consequentialismsep}.

Most of the above theories of good require that a moral reasoner understand complex features about
individuals' preferences, desires, or sensations in order to evaluate a moral action, making automated
consequentialist ethics difficult. Regardless of the theory of the good, a consequentialist ethical 
reasoner needs to evaluate a state of affairs, which encompass each involved individual's pleasure, 
preferences, welfare, freedom, rights, or whatever other criteria make a state good. This requires
judgements about whether or not a state of affairs actually satisifes the relevant criteria for goodness. 
These judgements are difficult and debateable, and any consequentialist decision requires many of 
these judgements for each individual involved. As systems become more complex and involve more people and 
more acts, making these judgements quickly becomes difficult, posing a scaling challenge for a 
consequentialist ethical reasoner. Perfect knowledge of tens of thousands of people's pleasure or 
preferences or welfare or rights is impossible. Either a human being 
assigns values to states of affairs, which quickly becomes difficult to scale, or the machine does, 
which requires massive common-sense, increases room for doubting the system's judgements, and simplifies
the judgements. This is a tractable problem, but is much more difficult than the Kantian task of formulating
and evaluating a maxim.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Aggregation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Once an automated consequentialist agent assigns a goodness measurement to each person in a state of affairs, it 
must also calculate an overall goodness measurement for the state of affairs. One approach to assigning
this value is to aggregate each person's individual goodness score into one complete score for a state. 
For example, under a simple welfare model, each person is assigned a welfare score and the total 
score for a state of affairs is the sum of the welfare scores for each involved person.
The more complex the theory of the good, the more difficult this aggregation becomes. For example, 
pluralistic theories struggle to explain how different kinds of value can be compared \cite{consequentialismsep}. 
How do we compare one unit of beauty to one unit of pleasure? Subjective theories of the good, such 
as those focused on the sensation of pleasure or an individual's preferences, present difficulties in 
comparing different people's subjective measures. Resolving this debate requires that the automated reasoner 
choose one specific aggregation algorithm, but those who disagree with this choice will not trust 
the reasoner's moral judgements. Moreover, for complex theories of the good, this aggregation algorithm
may be complex and may require a lot of data.

To solve this problem, some consequentialists reject aggregation entirely and instead prefer wholistic
evaluations of a state of affairs. While this approach no longer requires that a reasoner define an 
aggregation algorithm, the reasoner still needs to calculate a goodness measurement for a state of 
affairs. Whereas before the reasoner could restrict analysis to a single person, the algorithm must now 
evaluate an entire state wholistically. Evaluating the goodness of an entire state of affairs is more complicated
than evaluating the goodness of a single person. As consequentialists modulate between aggregation 
and wholistic evaluation, they face a tradeoff between the difficulty of aggregation and the complexity 
of goodness measurements for large states of affairs. This tradeoff also holds for an automated
consequentialist moral agent. Such an agent either needs to define an aggregation function, thus opening 
the door to those who disagree with this definition, or needs to evaluate the goodness of entire states
of affairs, which is a complex and data-intensive philosophical and technical challenge.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Prior Attempts to Formalize Consequentialism}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
None of the challenges described above are intractable or capture the full literature of 
all variations of consequentialism. I do not argue that consequentialism is 
impossible to automate. Instead, each of the challenges above requires that the developer 
``plant certain flags" and take a stance on certain philosophical debates. Such debates are present in 
any ethical theory, but consequentialism has more such points of difficulty than Kantian ethics and 
is thus more difficult to automate.

Because of its intuitive appeal, automated ethicists have tried to formalize consequentialism in the past.
These efforts cannot escape the debates outlines above. For example, Abel et. al represent ethics as a

@@

\cite[3]{util1}. While this is a convenient representation, it either leaves unanswered or 

@@

reward is accumulated in an MDP. It leaves open the question of what the reward function is and thus 

@@

and pain as bad.} as the theory of the good \cite[2]{util2}. Again, their proposal assumes that pleasure and pain can be 

@@

``makes its possessor good" \cite{vesep}. For example, under Aristotlean virtue ethics, virtues 

@@

in terms of the characteristic activity of human beings (in ethical terms, not telelological terms) \cite{snow}. 

@@

theory of the virtues which presents and justifices a list of the virtues. Such theories vary

@@

equanimity \cite{aristotle, mcrae}. Another theory is Sen's conception of the virtues as capabilites that create
``effective opportunities to undertake the actions and activities" an agent wants to engage in \cite{robeyns}. 

@@

itself up to criticism from those who disagree with this theory of the virtues. Unlike Kantian ethicists, 
who generally agree on the meaning of the FUL, virtue ethicists robustly debate which character traits qualify as virtues.%

@@

theory is often a person's entire moral character. While Kantians evaluate the maxim of an act and utilitarians

@@

require complex representations, an agent's ethical character and disposition is even more difficult
to represent. Consequentialism posed a data-collection problem in evaluating and representing states

@@

progress, whereas Kantian ethics immediately presents a formal rule to implement.%

@@

moral habit, which seems to suggest a machine learning approach. Artistotle, for example, argued 

>>>>

%
\begin{isabellebody}%
\setisabellecontext{whykant}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Choice of Ethical Theory%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this thesis, I automate Kantian ethics. In 2006, Powers posited that deontological theories are 
attractive candidates for automation because rules are generally computationally tractable \cite[1]{powers}. 
Intuitively, algorithms are rules or procedures for problem solving and deontology offers one such 
procedure for the problem of making ethical judgements. I will make this intuition precise by
arguing that deontological ethics is natural to formalize because rules generally require little additional
data about the world and are usually easy to represent to a computer. All ethical traditions have debates that an 
automated ethical system will need to take a stance on, but these debates are less frequent and controversial
for deontological ethics than for consequentialism and virtue ethics.

I do not aim to show that deontology is the only tractable theory to automate or
to present a comprehensive overview of all consequentialist or virtue ethical theories. Instead, I 
present a sample of some approaches in each tradition and argue that deontology is more straightforward 
to formalize than these approaches. Future work could and should address the challenges I outline in 
this section. The more ethical theories that computational tools can handle, the more valuable 
computational philosophy becomes both for philosophers and for AI agents. Insofar as my project serves 
as an early proof-of-concept for computational ethics, I choose to automate an ethical theory that 
poses fewer challenges than others.

I first present deontological ethics, then consequentialism, and finally virtue ethics. For each 
tradition, I present a crash course for non-philosophers and then explain some obstacles to automation, 
arguing that these obstacles are weakest in the case of deontology. Finally, I will present the specific 
deontological theory I am automating (Kantian ethics) and will argue that it is comparatively easier
to formalize. I will also outline the specific debates in the literature that my formalization takes
a stance on and potential challenges for formalizing deontology.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Deontological Ethics%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
\textbf{Crash Course on Deontology}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Deontological ethical theories evaluate actions as permissible, obligatory, or prohibited. The 
deontological tradition argues that an action should not be judged on its consequences, 
but rather on ``its confirmity with a moral norm" \citep{sepdeont}. In other words, deontological theories
define a set of moral norms or rules and evaluate actions based on their confirmity or lackthereof to
these rules. Deontologists do not believe that an agent should maximize the number of times that they 
conform to such rules; rather they argue that an agent should never violate any of the moral laws. A wrong
choice is wrong, regardless of its consequences.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Formalizing Deontology}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Deontology is immediately an attractive candidate for formalization because computers tend to 
understand rules; programming languages are designed to teach computers algorithms. Deontological
ethical theories give inviolable rules that an automated agent can apply, without having to adjust
the rules in changing situations or contexts. Moreover, because deontological theories focus on
the action itself, they require relatively little data. A deontological moral judgement does not require
as much information about context, consequences, or moral character as the other theories presented later in this
section. All that matters is the action and some limited set of circumstances in which it is performed.
I will later argue that, in the case of the specific deontological ethic that I implement (Kantian ethics), 
the action has a very ``thin" representation that is space and memory efficient and requires little data
about the external world.

Like all ethical traditions, deontology has debates that any implementation of automated deontological
ethics would need to resolve. Deontologists disagree about whether ethics should focus on agents and 
prescribe them action or should focus on the patients or potential victims of actions and their rights.
 Different deontological theories have different conceptions of what an action is, 
from the actual physical act to the agent's mental state at the time of acting to the principle upon
which the agent acted.

While these debates are certainly open, ``if any philosopher is regarded as central
to deontological moral theories, it is surely Immanuel Kant" \citep{sepdeont}. Out of the three ethical
traditions considered in this section, deontology has the most central representative in the form of 
Kant. Many modern deontologists claim to interpret Kant in a particular way, and thus agree on 
his ethic as the foundation of their theory but disagree in how to interpret it. In this paper, I will 
formalize Kantian ethics. Deontology's comparatively greater focus on Kant means that the choice of 
Kant as a guiding figure will be less controversial to deontologists than, for example, the choice of Bentham
as the guiding figure of consequentialism. Moreover, at the end of this section, I also argue that 
internal debates in the part of Kantian ethics that I focus on tend to be less controversial than those
in the consequentialist or virtue ethical traditions.

I do not argue that deontology is the only tractable theory to formalize, but instead that it is 
easier to formalize because it requires less data, is easily representable to a computer, and has fewer
and less controversial open debates than consequentialism or virtue ethics. That being said, any ethical
tradition has debates that an automated agent will need to take a stance on and deontological ethics
is no exception. Those who disagree with my stances (e.g. non-Kantians) will not trust my system's
judgements. This project is not irrelevant for them because it still serves as a case study in 
the power of computational ethics, but Kantians will trust my system's moral judgements the most.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Consequentialism%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
A consequentialist ethical theory is, broadly speaking, any ethical theory that evaluates an action by evaluating 
its consequences.\footnote{There is long debate about what exactly makes an ethical theory consequentialist \citep{consequentialismsep}. 
For this paper, I will focus on theories that place the moral worth of an act in its the consequences.} For example, 
utilitarianism is a form of consequentialism in which the moral action 
is the action that results in the best consequences or produces the most good \citep{utilsep}. The focus
on the consequences of action distinguishes consequentialists from deontologists, who derive the moral worth
of an action from the action itself. Some debates in the consequentialist tradition include 
which consequences of an action matter, what exactly constitutes a ``good" consequence, and how we can 
aggregate the consequences of an action over all the individuals involved.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Which Consequences Matter}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Because consequentialism evaluates the state of affairs following an action, this kind of ethical 
reasoning requires more knowledge
about the state of the world than deontology. Under a naive version of consequentialism, evaluating 
an action requires perfect knowledge of all consequences following an action. This requires that an 
automated ethical system somehow collect all of the infinite consquences following an action, a 
difficult, if not impossible, task. Moreover, compiling this database of consequences requires 
answering difficult questions about which consequences were actually caused by an action.\footnote
{maybe cite the debate about difficulties in determining causation?}

These challenges also apply to human reasoners, so most consequentialists do not actually adopt the naive
view that agents need to calculate all the consequences of their actions. Plausible strategies to avoid 
this problem include stopping calculation early because constant calculation paralyzes action or only 
evaluating consequences that the agent could reasonably forsee before acting. Another solution is the 
``proximate cause" approach, which only holds the agent responsible for the immediate consequences of their 
acts, but not for consequences resulting from others' voluntary responses to the agent's original act \citep{consequentialismsep}.

Even without understanding the details of these views, it is clear that they require 
more data than deontology and scale poorly with the complexity of the act being evaluated. Even 
if we cut off the chain of causal reasoning at some point based on 
one of the rules above, evaluating the consequences of an action is still data-intensive. Even evaluating
the first or immediate cause of an action requires knowledge about the state of the world before
and after an action, in addition to knowledge about the action itself. Consequentialism requires 
knowledge about 
the situation in which the act is performed and following the act, whereas deontology mostly requires 
knowledge about the act itself. For simple acts, collecting this data may not seem unreasonable, but as acts become
more complex and affect more people, the computational time and space required to calculate and store
their consequences increases. Collecting this data is theoretically possible, but is labor and resource
intensive. Deontology, on the other hand, does not suffer this scaling
challenge because acts that affect 1 person and acts that affect 1 million people share the same
representation.

The fact that consequentialism requires more knowledge about the world makes it more difficult to formalize.
Automated consequentialist ethical systems would need to represent complex states of the world and causal chains
in an efficient manner and reason about them. This both presents a difficult technical challenge and
impedes the usability of such a system. Such a system would need to come equipped with 
a large enough database of knowledge about the world to extrapolate the consequences of an actions and
up-to-date information about the state of the world at the moment of action. Not only does collecting 
and representing this data pose a technical challenge, it also creates a larger ``trusted code base" for 
the automated system. Trusting my deontological ethical reasoner merely requires trusting the logical implementation
of the categorical imperative and the formulation of a maxim. Trusting a consequentialist ethical 
reasoner, on the other hand, requires trusting both the logical machinery that actually evaluates the act and the 
background/situational knowledge that serves as an input to this machinery.

The challenge of understanding and representing the circumstances of action is not unique to consequentialism,
but is particularly acute for consequentialism. Deontologists robustly debate which circumstances 
of an action are ``morally relevant" and should be included in the formulation of the action.\footnote{ 
\cite{powers} identifies this as a challenge for automating Kantian ethics and briefly sketches 
solutions from \citet{constofreason}, \citet{silber}, and \citet{rawlsconstructivism}. }
Those using my system will need to use common-sense reasoning to determine which aspects of the circumstances
in which the action is performed are morally relevant and should thus be represented to the computer.\footnote{For more on the challenge of 
parsing of ethical dilemmas into maxims, see Section AI Ethics.} However, because deontology
merely evaluates the action itself, the surface of this debate is much smaller
than the debate about circumstances and consequences in a consequentialist system. An automated 
consequentialist system must make judgements about the act itself, the circumstances in which 
it is performed, and the circumstances following the act. The ``trusted code base" is 
smaller for deontology than for consequentialism. All ethical theories will require some set of 
circumstances and common-sense knowledge as part of the trusted code base, but this set is larger for 
consequentialism than for deontology.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Theory of the Good}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Another debate that an automated consequentialist reasoner would need to take a stance on is
the question of what qualifies as a ``good consequence," or what the theory of the good is. Hedonists associate
good with the presence of pleasure and the absence of pain. Preference utiliarians believe that good is 
the satisfaction of desires and is thus derived from individuals' preferences, as opposed to some
sensation of pleasure or pain. Other consequentialists adopt a pluralistic theory of value, under which 
many different kinds of things are good for different reasons. For example, Moore values beauty and truth 
and other pluralists value justice, love, and freedom \citep{moorepe}. Welfare utilitarians value a person's 
welfare and utilitarians of right value states of affairs in which respect for some set of
rights is maximized \citep{consequentialismsep}.

Most of the above theories of good require that a moral reasoner understand complex features about
individuals' preferences, desires, or sensations in order to evaluate a moral action, making automated
consequentialist ethics difficult. Regardless of the theory of the good, a consequentialist ethical 
reasoner needs to evaluate a state of affairs, which encompass each involved individual's pleasure, 
preferences, welfare, freedom, rights, or whatever other criteria make a state good. This requires
judgements about whether or not a state of affairs actually satisifes the relevant criteria for goodness. 
These judgements are difficult and debateable, and any consequentialist decision requires many of 
these judgements for each individual involved. As systems become more complex and involve more people and 
more acts, making these judgements quickly becomes difficult, posing a scaling challenge for a 
consequentialist ethical reasoner. Perfect knowledge of tens of thousands of people's pleasure or 
preferences or welfare or rights is impossible. Either a human being 
assigns values to states of affairs, which quickly becomes difficult to scale, or the machine does, 
which requires massive common-sense, increases room for doubting the system's judgements, and simplifies
the judgements. This is a tractable problem, but it is much more difficult than the deontological task of formulating
and evaluating an action.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Aggregation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Once an automated consequentialist agent assigns a goodness measurement to each person in a state of affairs, it 
must also calculate an overall goodness measurement for the state of affairs. One approach to assigning
this value is to aggregate each person's individual goodness score into one complete score for a state. 
For example, under a simple welfare model, each person is assigned a welfare score and the total 
score for a state of affairs is the sum of the welfare scores for each involved person.
The more complex the theory of the good, the more difficult this aggregation becomes. For example, 
pluralistic theories struggle to explain how different kinds of value can be compared \citep{consequentialismsep}. 
How do we compare one unit of beauty to one unit of pleasure? Subjective theories of the good, such 
as those focused on the sensation of pleasure or an individual's preferences, present difficulties in 
comparing different people's subjective measures. Resolving this debate requires that the automated reasoner 
choose one specific aggregation algorithm, but those who disagree with this choice will not trust 
the reasoner's moral judgements. Moreover, for complex theories of the good, this aggregation algorithm
may be complex and may require a lot of data.

To solve this problem, some consequentialists reject aggregation entirely and instead prefer wholistic
evaluations of a state of affairs. While this approach no longer requires that a reasoner define an 
aggregation algorithm, the reasoner still needs to calculate a goodness measurement for a state of 
affairs. Whereas before the reasoner could restrict analysis to a single person, the algorithm must now 
evaluate an entire state wholistically. Evaluating the goodness of an entire state of affairs is more complicated
than evaluating the goodness of a single person. As consequentialists modulate between aggregation 
and wholistic evaluation, they face a tradeoff between the difficulty of aggregation and the complexity 
of goodness measurements for large states of affairs. This tradeoff also holds for an automated
consequentialist moral agent. Such an agent either needs to define an aggregation function, thus opening 
the door to critique from those who disagree with this definition, or needs to evaluate the goodness of entire states
of affairs, which is a complex and data-intensive philosophical and technical challenge.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\textbf{Prior Attempts to Formalize Consequentialism}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
None of the challenges described above are intractable or capture the full literature of 
all variations of consequentialism. Instead, the challenges above require that the developer 
``plant certain flags" and take a stance on certain philosophical debates. Such debates are present in 
any ethical theory, but consequentialism has more such points of difficulty than deontology and 
is thus more difficult to automate.

Because of its intuitive appeal, computer scientists have tried to formalize consequentialism in the past.
These efforts cannot escape the debates outlined above. For example, Abel et al. represent ethics as a

@@

\citep[3]{util1}. While this is a convenient representation, it either leaves unanswered or 

@@

reward is accumulated in an MDP.\footnote{Generally, reward for an MDP is accumulated according to a 
``discount factor" $\gamma < 1$, such that if $r_i$ is the reward at time $i$, the total reward is $\sum_{i=0}^{\infty}\gamma^i r_i$.} 
It leaves open the question of what the reward function is and thus 

@@

and pain as bad.} as the theory of the good \citep[2]{util2}. Again, their proposal assumes that pleasure and pain can be 

@@

``makes its possessor good" \citep{vesep}. For example, under Aristotlean virtue ethics, virtues 

@@

in terms of the characteristic activity of human beings (in ethical terms, not telelological terms) \citep{snow}. 

@@

theory of the virtues which presents and justifies a list of the virtues. Such theories vary

@@

equanimity \citep{aristotle, mcrae}. Another theory is Sen's conception of the virtues as capabilites that create
``effective opportunities to undertake the actions and activities" an agent wants to engage in \citep{robeyns}. 

@@

itself up to criticism from those who disagree with this theory of the virtues. Any automated virtue
ethical agent will need to justify its choice of virtues.%

@@

theory is often a person's entire moral character. While deontologists evaluate the act itself and utilitarians

@@

require complex representations, an agent's ethical character and disposition are even more difficult
to represent to a computer. Consequentialism posed a data-collection problem in evaluating and representing states

@@

progress, whereas deontology immediately presents a formal rule to implement.%

@@

moral habit, which seems to be amenable to a machine learning approach. Artistotle, for example, argued 
